Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-03-18 ltx/BrainIDE.tex
Changes to the noise definition

\section{Generating Data}
To generate iEEG/LFP data we proceeded with the assumption of spatial homogeneity for the connectivity and activation function parameters, implying $w(r,r') = w\left(|r-r'|\right)$, $\nu_0(r) = \nu_0$ and $v_0(r) = v_0$. In addition we assume the synaptic transmission delay is negligible. The model was discretized in time  (see Appendix~\ref{Time Discretization}) and a stochastic term was added to drive the system and account for uncertainty in the model giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I_t\left(r\right) + e_t\left(r\right)
\end{equation}
where $\delta$ is the time step, $\lambda = 1-\alpha\delta$ and $e_t(r) \sim \mathcal{GP}(0,\Sigma)$. The IDE formulation in Eq.~\ref{DiscreteTimeModel} was solved using a numerical method described in Appendix~\ref{Space Discretization}. To generate iEEG/LFP data we used the output function 
\begin{equation}
	y_t(r_n) = \int_{\Omega}{m\left(|r_n-r'|\right)v_t\left(r'\right)dr'} + \varepsilon_t
\end{equation}
where $r_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\varepsilon_t \sim \mathcal{N}\left(0,\Sigma_y\right)$. The output kernel $m(|r-r'|)$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(|r-r'|\right) = \exp{\left(-\frac{(r-r')^\top(r-r')}{\sigma_m^2}\right)}.
\end{equation}

@@

\section{Derivation of Finite Dimensional State-Space Model}
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
	v_t\left(r\right) \approx \phi ^T\left(r\right) x_t
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $x_t$. As described in Eq.~\ref{Connectivity Kernel} the connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	\tilde{w}( |r-r'| ) = \theta^T\psi\left(|r-r'|\right)
\end{equation}
We will assume we know the parametric form of the connectivity basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get
\begin{equation}\label{reduced continuous model}
	\phi^T\left( r \right)x_{t+1} = \delta\theta^T\int_\Omega  {\psi \left(|r-r'|\right)f\left( \phi ^T\left(r'\right)x_t \right)dr'}  - \lambda\phi^T\left(r\right)x_t + \delta I_t\left(r\right) + e_t\left(r\right).
\end{equation}
By defining the term
\begin{equation}\label{DefGamma}
	\Gamma  = \int_\Omega  {\phi \left(r\right)\phi ^T\left(r\right)dr}
\end{equation}
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model
\begin{eqnarray}\label{Homogeneous SS Model}
	x_t &=& q(x_t,\theta,\lambda) + k(I_t) + \tilde{e}_t \\
	q(x_t,\theta,\lambda) &=& \delta\Gamma^{ - 1}\int_\Omega {\phi\left(r\right) \int_\Omega {f\left(\phi^T\left(r'\right)x_t\right)\psi^T \left(|r-r'|\right)dr'} dr} \theta - \lambda x_t \\
	k(I_t) &=& \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} \\
	\tilde{e}_t &=& \Gamma^{-1}\int_\Omega{\phi\left(r\right)e_t\left(r\right)dr} \\
\end{eqnarray}
where $\tilde{e}_t \sim \mathcal{N}\left(0,R' \Sigma R\right)$ and $R$ is the Cholesky decomposition of $\Gamma^{-1}$. The observation equation of the reduced model is
\begin{eqnarray}\label{ObservationEquation}
    y_t &=& z\left(x_t\right) + \varepsilon_t \\
	z\left(x_t\right) &=& \int_\Omega m\left(|r_n-r|\right)\phi^T\left(r\right) dr x_t.
\end{eqnarray}

@@

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{Colored noise}
We have
\begin{equation}
v_{t + 1}\left( \mathbf{r} \right)  = T_s\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( \bf{r'} \right)} \right)d{\bf{r'}}} + \lambda v_t\left( {\bf{r}} \right) + T_se_t\left( {\mathbf{r}} \right).
\end{equation}
where the disturbance $e_t(\mathbf r)$ is a zero-mean normally distributed noise process, spatially coloured but temporally white, with covariance

\begin{equation}
cov(e_t(\mathbf{s}),e_{t+\tau}(\mathbf{r}))=
\begin{cases}
\beta\left(\mathbf{s}-\mathbf{r}\right), & \tau=0 \\
0 & \mathrm{otherwise}
\end{cases}
\label{eq:FieldCovariance}
\end{equation}
The final state-space model is given by
\begin{equation}
x_{t + 1} = T_sq(x_t)\theta -\lambda x_t + T_s\Gamma^{-1}\int_\Omega  {\phi ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
then the covariance of the noise term is given by
\begin{eqnarray}
\mathbf \Sigma_w&{}={}&T_s^2 \times \mathbf{\Gamma}^{-1}\mathbf E[\int_{\mathcal{S}}\boldsymbol\phi\left(\mathbf s\right)e_t\left(\mathbf s\right)d\mathbf s\int_{\mathcal{S}}\boldsymbol\phi\left(\mathbf r\right)^{\top}e_t\left(\mathbf r\right)d\mathbf r]\mathbf{\Gamma}_{x}^{- \top} \nonumber \\
&=&T_s^2 \times\mathbf{\Gamma}^{-1}\iint\limits_{\mathcal{S}}\boldsymbol\phi\left(\mathbf s\right) \mathbf E[e_t\left(\mathbf s\right)e_t\left(\mathbf r\right)]\boldsymbol\phi\left(\mathbf r\right)^{\top}d\mathbf r d\mathbf s\mathbf{\Gamma}^{- \top} \nonumber\\
&=&{}T_s^2 \times\mathbf{\Gamma}^{-1}\iint\limits_{\mathcal{S}}\boldsymbol\phi\left(\mathbf s\right) \beta\left(\mathbf s- \mathbf r \right)\boldsymbol\phi\left(\mathbf r\right)^{\top}d\mathbf r d\mathbf s\mathbf{\Gamma}^{- \top} 
\end{eqnarray}

>>>>

\section{Generating Data}
To generate iEEG/LFP data we proceeded with the assumption of spatial homogeneity for the connectivity and activation function parameters, implying $w(r,r') = w\left(|r-r'|\right)$, $\nu_0(r) = \nu_0$ and $v_0(r) = v_0$. In addition we assume the synaptic transmission delay is negligible. The model was discretized in time  (see Appendix~\ref{Time Discretization}) and a stochastic term was added to drive the system and account for uncertainty in the model giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + \delta I_t\left(r\right) + e_t\left(r\right)
\end{equation}
where $\delta$ is the time step, $\lambda = 1-\alpha\delta$ and $e_t(r) \sim \mathcal{GP}(0,\Sigma)$\begin{todo}{
is a spatially colored but temporally white with covariance
\begin{equation}
 cov(e_t(\mathbf{r}),e_{t+\tau}(\mathbf{r'}))=
\begin{cases}
\beta\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
0 & \mathrm{otherwise}
\end{cases}
\label{eq:FieldCovariance}
\end{equation}}
\end{todo}. The IDE formulation in Eq.~\ref{DiscreteTimeModel} was solved using a numerical method described in Appendix~\ref{Space Discretization}. To generate iEEG/LFP data we used the output function 
\begin{equation}
	y_t(r_n) = \int_{\Omega}{m\left(|r_n-r'|\right)v_t\left(r'\right)dr'} + \varepsilon_t
\end{equation}
where $r_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\varepsilon_t \sim \mathcal{N}\left(0,\Sigma_y\right)$. The output kernel $m(|r-r'|)$ governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(|r-r'|\right) = \exp{\left(-\frac{(r-r')^\top(r-r')}{\sigma_m^2}\right)}.
\end{equation}

@@

\section{Derivation of Finite Dimensional State-Space Model}
In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
	v_t\left(r\right) \approx \phi ^T\left(r\right) x_t
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $x_t$. As described in Eq.~\ref{Connectivity Kernel} the connectivity kernel can also be considered a decomposition into a set of basis functions in 
\begin{equation}\label{DefKernelDecomp}
	\tilde{w}( |r-r'| ) = \theta^T\psi\left(|r-r'|\right)
\end{equation}
We will assume we know the parametric form of the connectivity basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get
\begin{equation}\label{reduced continuous model}
	\phi^T\left( r \right)x_{t+1} = \delta\theta^T\int_\Omega  {\psi \left(|r-r'|\right)f\left( \phi ^T\left(r'\right)x_t \right)dr'}  - \lambda\phi^T\left(r\right)x_t + \delta I_t\left(r\right) + e_t\left(r\right).
\end{equation}
By defining the term
\begin{equation}\label{DefGamma}
	\Gamma  = \int_\Omega  {\phi \left(r\right)\phi ^T\left(r\right)dr}
\end{equation}
and manipulating the model (see Appendix~\ref{Simplifying Decomposition}) we isolate the state vector to form a state-space model
\begin{eqnarray}\label{Homogeneous SS Model}
	x_t &=& q(x_t,\theta,\lambda) + k(I_t) + \tilde{e}_t \\
	q(x_t,\theta,\lambda) &=& \delta\Gamma^{ - 1}\int_\Omega {\phi\left(r\right) \int_\Omega {f\left(\phi^T\left(r'\right)x_t\right)\psi^T \left(|r-r'|\right)dr'} dr} \theta - \lambda x_t \\
	k(I_t) &=& \Gamma^{-1}\delta \int_\Omega{\phi \left(r\right) I_t\left(r\right)dr} \\
	\tilde{e}_t &=& \Gamma^{-1}\int_\Omega{\phi\left(r\right)e_t\left(r\right)dr} \\
\end{eqnarray}
\begin{todo}{where $\tilde{e}_t$ is a zero-mean normal white noise term with the covariance (see Appendix D)
\begin{equation} 
\mathbf \Sigma_w={}\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol\phi\left(\mathbf r\right) \beta\left(\mathbf r- \mathbf r' \right)\boldsymbol\phi\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} 
\end{equation}}

\end{todo}

where $\tilde{e}_t \sim \mathcal{N}\left(0,R' \Sigma R\right)$ and $R$ is the Cholesky decomposition of $\Gamma^{-1}$. The observation equation of the reduced model is
\begin{eqnarray}\label{ObservationEquation}
    y_t &=& z\left(x_t\right) + \varepsilon_t \\
	z\left(x_t\right) &=& \int_\Omega m\left(|r_n-r|\right)\phi^T\left(r\right) dr x_t.
\end{eqnarray}

@@

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{Colored noise}
We have
\begin{equation}
v_{t + 1}\left( \mathbf{r} \right)  = \delta\int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( \bf{r'} \right)} \right)d{\bf{r'}}} + \lambda v_t\left( {\bf{r}} \right) +e_t\left( {\mathbf{r}} \right).
\end{equation}
where the disturbance $e_t(\mathbf r)$ is a zero-mean normally distributed noise process, spatially coloured but temporally white, with covariance

\begin{equation}
cov(e_t(\mathbf{r}),e_{t+\tau}(\mathbf{r'}))=
\begin{cases}
\beta\left(\mathbf{r}-\mathbf{r'}\right), & \tau=0 \\
0 & \mathrm{otherwise}
\end{cases}
\label{eq:FieldCovariance}
\end{equation}
The final state-space model is given by
\begin{equation}
x_{t + 1} = \delta q(x_t)\theta -\lambda x_t + \Gamma^{-1}\int_\Omega  {\phi ( \mathbf{r} )e_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
then the covariance of the noise term is given by
\begin{eqnarray}
\mathbf \Sigma_w&{}={}&\mathbf{\Gamma}^{-1}\mathbf E[\int_{\Omega}\boldsymbol\phi\left(\mathbf r\right)e_t\left(\mathbf r\right)d\mathbf r\int_{\Omega}\boldsymbol\phi\left(\mathbf r'\right)^{\top}e_t\left(\mathbf r'\right)d\mathbf r']\mathbf{\Gamma}^{- \top} \nonumber \\
&=&\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol\phi\left(\mathbf r\right) \mathbf E[e_t\left(\mathbf r\right)e_t\left(\mathbf r'\right)]\boldsymbol\phi\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} \nonumber\\
&=&{}\mathbf{\Gamma}^{-1}\iint\limits_{\Omega}\boldsymbol\phi\left(\mathbf r\right) \beta\left(\mathbf r- \mathbf r' \right)\boldsymbol\phi\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top} 
\end{eqnarray}
